\\Large Rising temperatures alter how and where mammals move

### 2025-10-01 writing/manuscript.Rmd
tightened discussion and added closing paragraph

\noindent \textbf{Running Title}: Temperature alters mammal movement behavior

@@

Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010]. While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more severe physiological damage in a shorter span of time [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024]. Mammals' homeothermy can cause them to have difficulty dissipating heat, particularly when ambient temperatures are near or above body temperature. Large-bodied mammals are most likely to alter their movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016; @alston_temperature_2020; @verzuh_behavioural_2023], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023]. Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' seasonal phenology, fitness, and life history [@botero_evolutionary_2015; @wells_life_2022; @newediuk_climate_2024; @mclellan_effect_2015; @pigeon_drivers_2016]. However, behavioral responses to heat vary substantially across species depending on body size, vagility, and activity times [@mccain_body_2014; @williams_range_2018]. For example, warming temperatures have allowed deer (*Odocoileus* spp.) to shift northward, which has led to increased density of wolves (*Canis lupus*) and higher predation on caribou (*Rangifer tarandus caribou*) [@latham_invading_2011; @barber_potential_2018; @dickie_habitat_2024]. Preparing for and responding to future changes will require a detailed understanding of not only how how and where mammals move throughout the thermal landscape but also the cascade of effects that such changes will have at the population and community levels [@cunningham_opportunity_2021].

@@

\noindent We have demonstrated that temperature is an important driver of how and where large mammals move, and that mammals alter their seasonal and daily cycles in movement behaviour in response to changes in temperature. However, predicting mammals' responses to climate change remains a challenge, as habitats warm rapidly and mammals are exposed to increasingly novel and extreme conditions [@ipcc_climate_2023; @diffenbaugh_changes_2013; @botero_evolutionary_2015]. We leveraged the flexibility and interpretability of HGAMs to estimate mammals' nonlinear responses to temperature without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018]. In particular, tensor product interaction terms were crucial for estimating smooth, nonlinear changes in daily and seasonal trends with temperature with data from multiple individuals. By allowing interaction terms to be smooth bivariate functions, we were able to quantify changes in movement behaviour without the need to discretize time into distinct windows (e.g., day/night and seasons). We were thus able to present results using relatively simple and interpretable surface plots (Figs. \@ref(fig:dist) and \@ref(fig:hrsfs)) that conveyed detailed and nuanced estimates of change. This framework was essential for estimating how mammals behave at extreme temperatures. Despite the recent increase in frequency of extreme weather events in western Canada [@zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018], such events remain relatively rare, so estimating mammals' nonlinear responses required the flexibility provided by GAMs [@wood_generalized_2017]. Additionally, hierarchical models allowed us to produce statistically sound estimates that accounted for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @jeltsch_need_2025; @silva_too_2025]. Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously, so we suggest choosing each smooth term's basis size (`k`) carefully. While penalized splines and REML can help avoid unruly smooth terms, the penalty terms can struggle to restrict model complexity when datasets are particularly large [@wood_generalized_2017; @wood_generalized_2017-1].

Our results suggest that climate change will cause mammals to alter their seasonal movement behaviour, but not all species will respond similarly. For example, hotter temperatures caused boreal caribou to travel longer distances in fall, winter, and spring but less in summer, mirroring the findings of @stien_congruent_2012, @leclerc_determinants_2021, and @lessard_local_2025. The increase in movement rates during the colder seasons are likely partly due to shallower snow depth that results in greater mobility [@pedersen_quantifying_2021], but rising temperatures during snowy seasons also increase the risk of rain-on-snow events, which limit forage availability and increase time spent foraging [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025]. In contrast, the reduced movement rates during warmer summers are likely mainly due to caribou's reduced movement rates and use of thermal refugia, given ungulates' elevated risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023]. Therefore, although we predict that boreal caribou's yearly average movement rates will remain approximately unchanged throughout the 21^st^ century, the projections do not show the changes in seasonal movement phenology. However, this is an important caveat, since the effects of extreme temperatures are exacerbated by phenological mismatches with seasonal photoperiod [@walker_global_2019], including the timing of molting and reproduction. Earlier growth seasons in boreal and arctic regions have resulted in lower calf birth and survival [@post_climate_2008], while the lengthening of the growth season has allowed moose (*Alces alces*) and deer to encroach on boreal caribou habitat and increase the density of coyotes (*Canis latrans*), cougars, and wolves [@barber_potential_2018; @demars_incorporating_2023], whose movement behavior also depends on ambient temperature. Thus, rising temperatures affect more than each individual species' movement behaviour; they have complex and interconnected effects on an entire community and its trophic structure.

Work by @cunningham_opportunity_2021 discusses the importance of adopting a community-level perspective when assessing the effects of climate change on animals' movement behavior and phenology. For example, our results show that wolves responded to temperature by altering their movement rates similarly to caribou. Consequently, as tempeatures warm and both species increase their movement rates in fall, winter, and spring, the resulting increase in encounter rates [@martinez-garcia_how_2020] will likely increase predation risk when caribou are pregnant or with young calves. Additionally, while both species saw reduced movement rates during hot summer days, boreal wolves' use of anthropogenic linear features (e.g., roads, seismic lines) may allow them to reduce movement costs while increasing the chances of encountering heat-stressed prey [@whittington_caribou_2011; @dickie_faster_2017; @dickie_resource_2022]. Furthermore, caribou that attempt to reduce predation risk from wolves by avoiding wolf habitat may still risk increasing predation pressure from bears [@leblond_caribou_2016] and other encroaching predators [@demars_incorporating_2023; @labadie_global_2023].

Considerations about changes in trophic interactions leads to another important caveat about the results we present. The estimated changes movement behaviour cannot be fully attributed to physiological responses to temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature. For example, mountain goats' lower movement rates at temperatures near 0$^\circ$C in winter may be in part due to increased movement difficulty, since melting snow and rain-on-snow events increase the energetic costs of movement [@white_population_2025] and encounter rates with predators [@sullender_defining_2023]. However, milder temperatures can also increase plant growth, which allows goats to spend more time foraging within and less time travelling across patches [@charnov_optimal_1976; @white_mountain_2025]. Therefore, figure \@ref(fig:dist) illustrates an estimate of the total effects of temperature on each species movement rates, but it does not explicitly account for energetic costs. Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs. Additionally, forest cover, elevation, and distance from water are often correlated, so the estimated effects should be interpreted carefully, and they may not be applicable outside our study areas. For example, both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not explicitly account for differences in forage availability at different altitudes.

Forage quality and availability is often a limiting factor for mammals' ability to adapt to warming temperatures by moving to higher altitudes, since plants are slower to arrive in and establish in new habitats than mammals, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013]. Consequently, mammals that move to higher elevations may be forced to spend more time foraging [@mezzini_how_2025], which will increase their energetic expenditure as well as their chances of encountering predators (including cougars and grizzly bears) or competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023]. If local peaks become unsuitable habitat, and moving to nearby higher peaks requires traveling across low-quality and high-risk habitat (Figure \@ref(fig:s-rss-cc)), habitat fragmentation may prevent movement across ranges [@white_mountain_2025]. Translocating animals to locations with higher peaks may help avoid such a scenario, but the success of relocations is species-dependent, and attempts are not always successful [@ranc_memory_2022; @white_mountain_2025].

Future work should develop models that account for covariates that are specific to the study species and area, such as forest age and disturbance [@demars_incorporating_2023], terrain ruggedness [@white_mountain_2017], anthropogenic linear features [@dickie_faster_2017], and growing human presence [@sih_evolution_2011; @rice_conspicuous_2022] and risk of human-wildlife conflict [@abrahms_climate_2023; @weststrate_how_2024]. In particular, forests in western Canada are highly dynamic over time [@zhang_half-century_2015], and fires size and burn severity have increased in western Canada in recent decades [@zurowski_summer_2023; @parisien_abrupt_2023]. Thus, while our use of a static raster of forest cover allowed for simpler models and climate change projections, future studies should account for mammals' responses to temporal forest dynamics, including the impacts of fire [@zurowski_summer_2023; @parisien_abrupt_2023], harvest [@lochhead_linking_2022] and other extractive industries [@dickie_faster_2017; @dickie_habitat_2024], as well as the subsequent forest regeneration. However, the inclusion of temporally dynamic variables will require that any future projections of species' behavior depend on temporally dynamic estimates of how such variables will change over time with changes in climate and anthropogenic activity.

Species' adaptability to change depends mainly on its exposure, sensitivity, and adaptability to changes, and growing evidence suggests that all three factors will limit mammals' ability to adapt to climate change [@ameca_y_juarez_natural_2012; @diffenbaugh_changes_2013]. Exposure to warming temperatures is widespread throughout the western Canada [@bunnell_global_2011; @kienzle_has_2018; @eum_potential_2023], and changes in temperature and seasonal phenology may be too rapid for many species to respond to [@diffenbaugh_changes_2013]. Species' adaptability is further challenged by the the difficulty in responding to many concurrent, multifarious, and interconnected changes [@diffenbaugh_changes_2013; @polazzo_measuring_2024]. Landscapes may thus become unable to support species and population sizes they supported historically, and species may begin selecting against historical ranges in the future [@williams_range_2018]. Therefore, it is crucial for conservationists to distinguish between a species' ideal environmental space and its best available geographic space [*sensu* @matthiopoulos_defining_2022]. Accelerating multifarious change in species' movement behaviour and energetic costs require conservation measures to be proactive and anticipate future changes, especially when future conditions are projected to be substantially different from historical conditions. @polazzo_measuring_2024 present a framework for estimating species' responses to multifarious change using GAMs that is particularly useful when estimating species' responses to interactions between environmental variables (e.g., responses to changes in both temperature and limiting nutrients). By accounting for the effects of multiple variables at once, they demonstrate how a species' response to one variable can depend on other concurrent changes. Estimating responses to multifarious change is crucial in assessing the effects of the effects of rising temperatures on community dynamics, including species' selection for current habitats as well as changes in the habitats' carrying capacities and trophic interactions. Protected areas must thus adapt to support populations not only in present conditions but also as species' needs and behavior change over the coming decades [@simmons_tough_2021; @livingstone_integrative_2023]. Recent evidence has questioned whether current protected areas will be effective in the long term [@loarie_velocity_2009; @diffenbaugh_changes_2013; @williams_range_2018], and increasingly more experts have been calling for proactive and long-term conservation. By identifying and actively protecting climate change refugia and corridors, we may be able to reduce the effects of climate change on mammalian movement behaviour and community-level responses and ensure that protected areas may be viable for decades to come [@stralberg_toward_2020; @hua_biodiversity_2022; @graham_prioritizing_2019; @michalak_distribution_2018].

>>>>

\noindent \textbf{Running Title}: Temperature alters mammal movement behaviour

@@

Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010]. While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more severe physiological damage in a shorter span of time [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024]. Mammals' homeothermy can cause them to have difficulty dissipating heat, particularly when ambient temperatures are near or above body temperature. Large-bodied mammals are most likely to alter their movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016; @alston_temperature_2020; @verzuh_behavioural_2023], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023]. Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' seasonal phenology, fitness, and life history [@botero_evolutionary_2015; @wells_life_2022; @newediuk_climate_2024; @mclellan_effect_2015; @pigeon_drivers_2016]. However, behavioural responses to heat vary substantially across species depending on body size, vagility, and activity times [@mccain_body_2014; @williams_range_2018]. For example, warming temperatures have allowed deer (*Odocoileus* spp.) to shift northward, which has led to increased density of wolves (*Canis lupus*) and higher predation on caribou (*Rangifer tarandus caribou*) [@latham_invading_2011; @barber_potential_2018; @dickie_habitat_2024]. Preparing for and responding to future changes will require a detailed understanding of not only how how and where mammals move throughout the thermal landscape but also the cascade of effects that such changes will have at the population and community levels [@cunningham_opportunity_2021].

@@

\noindent We have demonstrated that temperature is an important driver of how and where large mammals move, and that it affects mammals' seasonal and daily movement behaviour in complex and nonlinear ways. However, predicting mammals' responses to climate change remains a challenge, as habitats are warming rapidly and mammals are exposed to increasingly novel and extreme conditions [@ipcc_climate_2023; @diffenbaugh_changes_2013; @botero_evolutionary_2015]. We leveraged the flexibility and interpretability of HGAMs to estimate mammals' nonlinear responses to temperature without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018]. In particular, tensor product interaction terms were crucial for estimating smooth, nonlinear changes in daily and seasonal trends with temperature with data from multiple individuals. By allowing interaction terms to be smooth bivariate functions, we were able to quantify changes in movement behaviour without the need to discretize time into distinct windows (e.g., day/night and seasons). We were thus able to present results using relatively simple and interpretable surface plots (Figs. \@ref(fig:dist) and \@ref(fig:hrsfs)) that conveyed nuanced, continuous-time estimates of change. This framework was essential for estimating how mammals behave at extreme temperatures. In fact, despite the recent increase in frequency of extreme heat events in western Canada [@zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018], such events remain relatively rare, so estimating mammals' nonlinear responses required the flexibility provided by GAMs [@wood_generalized_2017]. Additionally, hierarchical models allowed us to produce statistically sound estimates that accounted for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @jeltsch_need_2025; @silva_too_2025]. Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously, so we suggest choosing each smooth term's basis size (`k`) carefully. While penalized splines and REML can help avoid unruly smooth terms, the penalty terms can struggle to restrict model complexity when datasets are particularly large [@wood_generalized_2017; @wood_generalized_2017-1].

Species varied greatly in how they adapted their movement behaviour to temperature. For boreal caribou, hotter temperatures increased movement rates in fall, winter, and spring but decreased them in summer, mirroring the findings of @stien_congruent_2012, @leclerc_determinants_2021, and @lessard_local_2025. The increase in movement rates during the colder seasons are likely partly due to shallower snow depth that results in greater mobility [@pedersen_quantifying_2021], but rising temperatures during snowy seasons also increase the risk of rain-on-snow events, which limit forage availability and increase time spent foraging [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025]. The reduced movement rates during warmer summers are likely mainly due to more frequent resting and use of thermal refugia, given ungulates' elevated risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023]. Therefore, although we predict that boreal caribou's yearly average movement rates will remain approximately unchanged throughout the 21^st^ century, the projections do not show the changes in seasonal movement phenology. This is an important caveat, since the effects of extreme temperatures are exacerbated by phenological mismatches with seasonal photoperiod [@walker_global_2019], including the timing of molting and reproduction. Earlier growth seasons in boreal and arctic regions have resulted in lower calf birth and survival [@post_climate_2008], while the lengthening of the growth season has allowed moose (*Alces alces*) and deer to encroach on boreal caribou habitat and increase the density of coyotes (*Canis latrans*), cougars, and wolves [@barber_potential_2018; @demars_incorporating_2023], whose movement behaviour also depends on ambient temperature. Boreal wolves responded to temperature similarly to boreal wolves, and habitat selection strength was not projected to change noticeably by 2100, but wolves were predicted to move less in future decades. In contrast, cougars showed markedly different seasonal responses to temperature, with relatively warmer temperatures always resulting in less travel, which resulted in strong predicted decreases in movement rates over the 21^st^ century. However, spatial heterogeneity in habitat, climate, and cougar physiology and behaviour may cause cougars in boreal regions to move differently from those in our study. Still, rising temperatures clearly affect more than each individual species' movement behaviour; they have complex and interconnected effects on entire communities and their trophic structures.

@cunningham_opportunity_2021 discuss the importance of adopting a community-level perspective when assessing the effects of climate change on animals' movement behaviour and phenology. For example, as temperatures warm and both boreal caribou and wolves increase their movement rates in fall, winter, and spring, they also increase their encounter rates [@martinez-garcia_how_2020], which will likely increase predation risk when caribou are pregnant or with young calves. Additionally, while both species saw reduced movement rates during hot summer days, boreal wolves' use of anthropogenic linear features (e.g., roads, seismic lines) may allow them to reduce the thermal costs of movement while increasing the chances of encountering heat-stressed prey [@whittington_caribou_2011; @dickie_faster_2017; @dickie_resource_2022]. Furthermore, caribou that attempt to reduce predation risk from wolves by avoiding wolf habitat may still risk increasing predation pressure from bears [@leblond_caribou_2016] and the aforementioned encroaching predators [@demars_incorporating_2023; @labadie_global_2023].

Considerations about changes in trophic interactions leads to another important caveat about the results we present. The estimated changes movement behaviour cannot be fully attributed to physiological responses to temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature. For example, mountain goats' lower movement rates at temperatures near 0$^\circ$C in winter may be in part due to increased movement difficulty, since melting snow and rain-on-snow events increase the energetic costs of movement [@white_population_2025] and encounter rates with predators [@sullender_defining_2023]. However, milder temperatures can also increase plant growth, which allows goats to spend more time foraging within and less time travelling across patches [@charnov_optimal_1976; @white_mountain_2025]. Therefore, figure \@ref(fig:dist) illustrates an estimate of the total effects of temperature on each species movement rates, but it does not explicitly account for energetic costs. Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs. Additionally, forest cover, elevation, and distance from water are often correlated, so the estimated effects should be interpreted carefully, and they may not be applicable outside each species' study area. For example, both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not explicitly account for differences in forage availability at different altitudes.

Climate change has also had extensive impacts on plant growth and phenology [@cleland_shifting_2007; @denny_performance_2019; @tysor_countergradient_2025] with consequences on herbivores' behaviour and fitness [@post_climate_2008; @aikens_greenscape_2017]. Forage quality and availability are often a limiting factor for herbivores' ability to escape hotter temperatures by moving to higher altitudes or latitudes, since plants are slower to arrive in and establish in new habitats, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013]. Consequently, mammals that move to higher elevations may be forced to spend more time foraging [@mezzini_how_2025], which will increase both their energetic expenditure and encounter rates with predators and competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023]. If local peaks become unsuitable habitat, and moving to nearby higher peaks requires traveling across low-quality and high-risk habitat (Figure \@ref(fig:s-rss-cc)), habitat fragmentation may prevent movement across ranges [@white_mountain_2025]. Translocating animals to locations with higher peaks may help avoid such a scenario, but the success of relocations is species-dependent, and attempts are not always successful [@ranc_memory_2022; @white_mountain_2025].

Growing evidence suggests that mammals' exposure, sensitivity, and adaptability to climate change will limit their ability to adapt to change [@ameca_y_juarez_natural_2012; @diffenbaugh_changes_2013]. Exposure to warming temperatures is widespread throughout western Canada [@bunnell_global_2011; @kienzle_has_2018; @eum_potential_2023], and changes in temperature and seasonal phenology may be too rapid for many species to respond to [@diffenbaugh_changes_2013]. Species' adaptability is further challenged by the difficulty in responding to concurrent, multifarious, and interconnected changes [@diffenbaugh_changes_2013; @polazzo_measuring_2024]. In the future, landscapes may thus become unable to support species and population sizes they supported historically, and species may begin selecting against historical ranges [@williams_range_2018]. Therefore, it is crucial for conservationists to distinguish between a species' ideal environmental space and its best available geographic space [*sensu* @matthiopoulos_defining_2022]. Accelerating multifarious change in species' movement behaviour and energetic costs require conservation measures that are proactive and anticipate future changes. @polazzo_measuring_2024 present a framework for estimating species' responses to multifarious change using GAMs that is particularly useful when estimating species' (nonlinear) responses to interactions between environmental variables (e.g., changes in both temperature and limiting nutrients). By accounting for the effects of multiple variables at once, they demonstrate how a species' response to one variable can depend on other concurrent changes. Estimating responses to multifarious change is crucial in assessing mammals' responses to hotter temperatures, altered phenology, and novel community dynamics. Protected areas must support populations not only in present conditions but also as species' needs and behaviour change over the decades [@simmons_tough_2021; @livingstone_integrative_2023]. Recent evidence has questioned whether current protected areas will be effective in the long term [@loarie_velocity_2009; @diffenbaugh_changes_2013; @williams_range_2018], and increasingly more experts have been calling for proactive and long-term conservation. By identifying and actively protecting climate change refugia and corridors, we may be able to reduce the effects of climate change on mammalian movement behaviour and community-level responses and ensure that protected areas may be viable for decades to come [@stralberg_toward_2020; @hua_biodiversity_2022; @graham_prioritizing_2019; @michalak_distribution_2018].

Future work should develop models that account for covariates that are specific to the study species and area, such as forest age and disturbance [@demars_incorporating_2023], terrain ruggedness [@white_mountain_2017], anthropogenic linear features [@dickie_faster_2017], and growing human presence in wild spaces [@sih_evolution_2011; @rice_conspicuous_2022], including the growing risk of human-wildlife conflict [@abrahms_climate_2023; @weststrate_how_2024]. In particular, forests in western Canada are highly dynamic over time [@zhang_half-century_2015], and fires size and burn severity have increased in western Canada in recent decades [@zurowski_summer_2023; @parisien_abrupt_2023]. Thus, while our use of a static raster of forest cover allowed for simpler models and climate change projections, future studies should account for mammals' responses to temporal forest dynamics, including the impacts of fire [@zurowski_summer_2023; @parisien_abrupt_2023], logging [@lochhead_linking_2022] and other extractive industries [@dickie_faster_2017; @dickie_habitat_2024]. However, the inclusion of temporally dynamic variables will require that any future projections of species' behaviour depend on temporally dynamic estimates of how such variables will change over time with changes in climate and anthropogenic activity.

We have demonstrated that temperature has complex and nonlinear effects on how and where mammals move. Rising temperatures have impacted mammals' daily and seasonal movement phenology, and changes are expected to intensify throughout the 21^st^ century. Quantifying how temperature affects mammalian behavior and fitness is a first step towards developing an understanding of the effects of climate change on landscape dynamics and the effectiveness of protected areas in the future. By leveraging the flexibility and statistical power of Hierarchical Generalized Additive Models, we have provided researchers with malleable and data-driven methods for assessing species' responses to climate change. We hope that future work will develop more species-specific models with the intent of developing a community-level framework for quantifying and predicting how communities will respond to climate change.
